---
title: Modeling and Predicting the Occurrence of Diabetes using Machine Learning Algorithm
  for Classification
author: ''
date: "`r Sys.Date()`"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment = NA, warning = FALSE, message = FALSE,
                      fig.height = 5, fig.width = 7)
```

## Results 
### Load the Required Libraries
```{r echo=FALSE}
library(ISLR2)
library(MASS)
library(caret)
library(splines)
library(pROC)
library(rattle)
library(recipes)
library(lava)
library(sjmisc)
library(igraph)
library(e1071)
library(hardhat)
library(ipred)
library(caret)
library(sjPlot)
library(nnet)
library(wakefield)
library(kknn)
library(dplyr)
library(nnet)
library(caTools)
library(ROCR)
library(stargazer)
library(dplyr)
library(nnet)
library(caTools)
library(ROCR)
library(stargazer)
library(ISLR)
library(ISLR2)
library(MASS)
library(splines)
library(splines2)
library(pROC)
library(ISLR)
library(ISLR2)
library(MASS)
library(splines)
library(splines2)
library(pROC)
library(randomForest)
library(rpart)
library(rpart.plot)
library(rattle)
library(ISLR2)
library(MASS)
library(splines)
library(pROC)
library(rattle)
library(rpart)
library(party)
library(partykit)
library(ggplot2)
library(tune)
library(TunePareto)
library(ISLR2)
library(MASS)
library(caret)
library(splines)
library(pROC)
library(rattle)
library(ggplot2)
library(devtools)
library(predict3d)
library(psych)
library(dplyr)
library(gtsummary)
library(DescTools)
library(nortest) 
library(lmtest)
library(sandwich)
library(sjmisc)
library(ggplot2)
library(stargazer)
```

### Load the Data
```{r}
diabetes_data <- read.csv("diabetes_prediction_dataset.csv")
attach(diabetes_data)
head(diabetes_data,5)
```

### Summary Statistics
```{r}
options(scipen = 999)
knitr::kable(
  describeBy(diabetes_data[c(-1,-5)]) %>% round(2) 
)
```


# Model Estimation
## Model One: Classification and Regression Tree (CART) Model
```{r}
diabetes_data$gender <- as.factor(diabetes_data$gender)
diabetes_data$smoking_history <- as.factor(diabetes_data$smoking_history)
diabetes_data$diabetes <- factor(diabetes_data$diabetes, levels = c(0,1),
                                 labels = c("No", "Yes"))
head(diabetes_data,10)
```

### Take a sample of 300 observations for easier code execusion
```{r}
set.seed(1234)
sample_size <- 300
random_indices <- sample(1:nrow(diabetes_data), size = sample_size, replace = FALSE)
sample <- diabetes_data[random_indices, ]
head(sample,10)
```

```{r}
set.seed(1234)
train_model <- trainControl(method = "repeatedcv", number = 5, repeats=10)

model.cart2 <- train( diabetes ~ gender + age + hypertension + heart_disease + smoking_history + bmi + HbA1c_level + blood_glucose_level,
  data = sample,
  method = "rpart",
  trControl = train_model)
```

### Model Summary
```{r}
model.cart2
```

### Confusion Matrix
```{r}
confusionMatrix(predict(model.cart2, sample), 
                reference=sample$diabetes, positive="Yes")
```


### Variable Importance
```{r}
varImp(model.cart2)
```

### Plot the Variable Importance
```{r}
plot(varImp(model.cart2))
```


## Model Two: Random Forest
```{r}
set.seed(1234)
model.rf2 <- train(
 diabetes ~ gender + age + hypertension + heart_disease + smoking_history + bmi + HbA1c_level + blood_glucose_level,
  data = sample,
  method = "rf",
  trControl = train_model)
model.rf2
```


### Confusion Matrix
```{r}
confusionMatrix(predict(model.rf2, sample), 
                reference=sample$diabetes, positive="Yes")
```



### Obtain variable importance
```{r}
varImp(model.rf2)
```

```{r}
plot(varImp(model.rf2), main = "Variable Importance for the Random Forest Model")
```

## Model Three: k-Nearest Neighbors
```{r}
set.seed(1234)
trControl <- trainControl(method = "repeatedcv",number = 5,repeats = 10)
model.knn2 <- train(diabetes ~ gender + age + hypertension + heart_disease + smoking_history + bmi + HbA1c_level + blood_glucose_level,
  data = sample,
  method = "knn",
  tuneLength = 5,
  trControl = trControl,
  preProc = c("center", "scale")
)
```

### View the Final Model
```{r}
model.knn2
```

### Classification Accuracy
```{r}
confusionMatrix(predict(model.knn2, sample), 
                reference=sample$diabetes, positive="Yes")
```


## Variable Importance
```{r}
varImp(model.knn2)
```


```{r}
plot(varImp(model.knn2), main = "Variable Importance for the K-NN Model")
```

## Model Four: Naive Bayes
```{r warning=FALSE}
set.seed(1234)
trControl <- trainControl(method = "repeatedcv",number = 5,repeats = 10)
model.nb2 <- train(diabetes ~ gender+ age + hypertension + heart_disease + smoking_history + bmi + HbA1c_level + blood_glucose_level,
  data = sample,
  method = "naive_bayes",
  tuneLength = 5,
  trControl = trControl,
  preProc = c("center", "scale")
)
```

### View the Model
```{r}
model.nb2
```

### Prediction and Classification Accuracy
```{r}
confusionMatrix(predict(model.nb2, sample), 
                reference=sample$diabetes, positive="Yes")
```


### Variable Importance
```{r}
varImp(model.nb2)
```


```{r}
plot(varImp(model.nb2), main = "Variable for the Naive Bayes Model")
```




## Model Five: Support Vector Machine (SVM)
```{r}
set.seed(1234)
trControl <- trainControl(method = "repeatedcv",number = 5,repeats = 10)
model.svm2 <- train(diabetes ~ gender + age + hypertension + heart_disease + smoking_history + bmi + HbA1c_level + blood_glucose_level,
  data = sample,
  method = "svmLinear",
  tuneLength = 5,
  trControl = trControl,
  preProc = c("center", "scale")
)
```

### View the Model
```{r}
model.svm2
```

### Prediction and Classification Accuracy
```{r}
confusionMatrix(predict(model.svm2, sample), 
                reference=sample$diabetes, positive="Yes")
```


### Variable Importance
```{r}
varImp(model.svm2)
plot(varImp(model.svm2), main = "Variable Importance for Support Vector Machine")
```


\newpage
## Reference
Mustafa, M. (2023). Diabetes prediction dataset. Kaggle.com. https://www.kaggle.com/datasets/iammustafatz/diabetes-prediction-dataset





